---
title: "Differential expression analysis of GSE29378 based on ADNI 1.5T 3Yr results from Elias"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

# source('/mnt/Github/knit2synapse-1/R/knitFile2Synapse.R')
# knitFile2Synapse(file = "./DiffExpAnal_GSE29378_MA.Rmd", owner = 'syn4185900', name = "Differential Expression Analysis GSE29378 MA",overwrite=F)

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

## Needs the dev branch
library(rGithubClient)

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R")
file.sources = sapply(file.sources,function(x){return(paste('../R/lib',x,sep='/'))})
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE)
```

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn3721177';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis based on image features';

thisFileName <- 'DiffExpAnal_GSE29378_MA.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Takeda_contract", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
```
### Image Analysis
1. Image features used:
  * FreeSurfer.thickness..median  
  * mean.curvature..median         
  * Volume
  * travel.depth..median  
  * FreeSurfer.convexity..sulc...median
```{r load.data}
adni.mem.1_5T.3Y <- read.csv('./adni.mem_ranked_1.5T_3Yr_takeda_4_6_15.csv')
label.name.conversion <- read.csv('./label_names_conversion.csv')

adni.mem.1_5T.3Y <- dplyr::mutate(adni.mem.1_5T.3Y, 
                                  sig.feature.all = a.ks.pvals <= 0.05 & a.cor.pvals <= 0.05,
                                  sig.feature.aks = a.ks.pvals <= 0.05,
                                  feature = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[-(length(x))],collapse='.')}),
                                  region = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[(length(x))],collapse='.')}))
adni.mem.1_5T.3Y <- merge(adni.mem.1_5T.3Y,label.name.conversion, by.x = 'region', by.y = 'Label')
sig.table.all <- xtabs(sig.feature.all ~ feature + region,data = adni.mem.1_5T.3Y)
colnames(sig.table.all) <- label.name.conversion[match(label.name.conversion$Label,colnames(sig.table.all)),1]

unique.features <- unique(sapply(str_split(rownames(sig.table.all),'\\.'),function(x){x <- x[x!=""];l <- length(x);ifelse(l>1,paste(x[-(l)],collapse='.'),x)}))

region.ranking <- colSums(sig.table.all[rownames(sig.table.all) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]

unique.regions.correlated <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

sig.table.aks <- xtabs(sig.feature.aks ~ feature + Label.name, data = adni.mem.1_5T.3Y)
region.ranking <- colSums(sig.table.aks[rownames(sig.table.aks) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]
unique.regions <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

```
2. Significantly changed regions are: `r unique.regions`
3. Significantly changed regions that are also correlated with ADNI.MEM are: `r unique.regions.correlated`
4. Data were obtained from GSE29378 (http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29378), and following brain regions were analysed for differential expression between case and control:
    * hippocampus CA1 and CA3
```{r download.data, cache=TRUE}
# Load series and platform data from GEO
GSET <- getGEO("GSE29378", GSEMatrix =TRUE)
ALL_USED_IDs <- 'ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE29nnn/GSE29378/matrix/'

# Log2 transformation (if needed)
EXPR <- exprs(GSET[[1]])
qx <- as.numeric(quantile(EXPR, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) || (qx[6]-qx[1] > 50 && qx[2] > 0) || (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
if (LogC) { 
  EXPR[which(EXPR <= 0)] <- NaN
  EXPR <- log2(EXPR)
}
tmp <- fData(GSET[[1]])[,"Symbol"]
EXPR <- EXPR[tmp!='',]
GNames <- droplevels(tmp[tmp!=''])

Collapsed.obj <- collapseRows(EXPR,GNames,rownames(EXPR))
EXPR <- Collapsed.obj$datETcollapsed
```

```{r covariates, cache=TRUE, fig.width=15, fig.height=10}
# Phenotype data
PHENO.DATA <- pData(phenoData(GSET[[1]]))
PHENO.DATA <- PHENO.DATA[, c('characteristics_ch1',paste('characteristics_ch1',1:11,sep='.'))]
PHENO.DATA <- dplyr::rename(PHENO.DATA, 
                            subjectnumber = characteristics_ch1, 
                            region = characteristics_ch1.1, 
                            gender = characteristics_ch1.2, 
                            age = characteristics_ch1.3, 
                            pmi = characteristics_ch1.4,
                            plaquescore = characteristics_ch1.5,
                            braakstage = characteristics_ch1.6,
                            apoe = characteristics_ch1.7,
                            diseaseduration = characteristics_ch1.8,
                            brainbank = characteristics_ch1.9,
                            chipnumber = characteristics_ch1.10,
                            locationchip = characteristics_ch1.11)
PHENO.DATA <- lapply(PHENO.DATA,function(cols){
  tmp <- sapply(cols,function(x){
    tmp <- strsplit(as.character(x),':')[[1]][2];
    tmp <- tolower(tmp);
    tmp <- str_trim(tmp, side = 'both');   
    tmp <- str_replace_all(tmp,"[[:punct:]]","_")
    tmp <- str_replace_all(tmp," ","_")
    tmp <- gsub('cingulate','singulate',tmp);
    return(tmp)
  })
  tmp <- factor(tmp)
})

# Convert numeric covariates to numbers
FactCovariates <- c("subjectnumber","region","gender","apoe","brainbank","chipnumber","locationchip","plaquescore","braakstage")   
ContCovariates <- c("age","pmi","diseaseduration")

PHENO.DATA[ContCovariates] <- lapply(PHENO.DATA[ContCovariates],
                                     function(x){as.numeric(gsub('_','.',as.character(x)))})
PHENO.DATA <- as.data.frame(PHENO.DATA)


PHENO.DATA <- dplyr::mutate(PHENO.DATA, status = ifelse(plaquescore %in% c(2,3),'AD','Control'))
PHENO.DATA$status <- factor(PHENO.DATA$status)

rownames(PHENO.DATA) <- colnames(EXPR)
```
Correlation between covariates
```{r pheno.correlation, fig.width=7,fig.height=7}
COVARIATES.CORRELATION = getAssociationStatistics(PHENO.DATA, PVAL = 1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

```{r model.fit, cache=TRUE, fig.height=10, fig.width=15}
# Find PC of gene expression and significant covariates that are highly correlated with PCs
preAdjustedSigCovars <- runPCAandPlotCorrelations(EXPR, PHENO.DATA,'expression', isKeyPlot=TRUE)

# Plot significant data
print(preAdjustedSigCovars$PC_res[[2]]$plotData)
print(paste('Significant covariates are:',paste(preAdjustedSigCovars$significantCovars,collapse=',')))

# Get design matrix
DESIGN <- getDesignMatrix(dplyr::select(PHENO.DATA,status,region, age, chipnumber), Intercept = F)
print(paste('Fitting a linear model with following coefficients:',paste(colnames(DESIGN$design),collapse=',')))
DESIGN$design <- DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Calculate correlation between donors
correlation <- duplicateCorrelation(EXPR, design=DESIGN$design, block=PHENO.DATA$subjectnumber)
print(paste('Subject effect is accounted as random effect in the model'))
print(paste('Consensus correlation is', correlation$consensus.correlation))

# Fit linear model
FIT <- lmFit(EXPR, design = DESIGN$design, block = PHENO.DATA$subjectnumber, correlation = correlation$consensus.correlation)

# Calculate residuals matrix
RESIDUAL <- residuals.MArrayLM(FIT,EXPR)

# Find correlation between residuals and covariates 
postAdjustedSigCovars <- runPCAandPlotCorrelations(RESIDUAL, PHENO.DATA, 'residual expression', isKeyPlot=TRUE)

if (length(postAdjustedSigCovars$significantCovars) != 0)
  print(paste('Following coefficients have to be included in the model:',paste(postAdjustedSigCovars$significantCovars,collapse=',')))

# Make contrast to check differential expression
CONT.NAMES <- colnames(FIT$coefficients)[grep('status',colnames(FIT$coefficients))]
CONT.NAMES <- combn(CONT.NAMES,2)
CONT.NAMES <- apply(CONT.NAMES,2,paste,collapse='-')
  
CONT <- makeContrasts(contrasts=CONT.NAMES,levels=colnames(FIT$coefficients))

# Refit contrasts
FIT.CONTRAST <- contrasts.fit(FIT,CONT)
  
# Estimate moderated t-statistics
FIT.CONTRAST <- eBayes(FIT.CONTRAST)
  
# Obtain all the differential expession combinations
DEXP <- list()
DEXP$logFC <- data.frame(row.names = rownames(EXPR))
DEXP$adj.P.Val <- data.frame(row.names = rownames(EXPR))
  
for (i in colnames(CONT)){
  tmp <- topTable(FIT.CONTRAST, coef=i, number=dim(EXPR)[1])    
  DEXP$logFC[,i] <- tmp[rownames(DEXP$logFC),'logFC']
  DEXP$adj.P.Val[,i] <- tmp[rownames(DEXP$adj.P.Val),'adj.P.Val']
}
  
# Get number of significantly differentialy expressed genes with adj.P.Val <= 0.05 and abs(logFC) >= 0
DEXP$SIG.EXP <- DEXP$adj.P.Val<=0.05 & abs(DEXP$logFC) >= 0   
DEXP$NUM.SIG.EXP <- colSums(DEXP$SIG.EXP)
```

```{r print.fc.pval}
# Print results
tmp <- as.data.frame(DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR <= 0.05'
kable(tmp)

# Get fold change and pvalues
FC <- DEXP$logFC
colnames(FC) <- 'GSE29378.MA.Hippocampus.AD-Control'
FC <- rownameToFirstColumn(DEXP$logFC,"hgnc_symbol")

PVAL<- DEXP$adj.P.Val
colnames(PVAL) <- 'GSE29378.MA.Hippocampus.AD-Control'
PVAL <- rownameToFirstColumn(DEXP$adj.P.Val,"hgnc_symbol")
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis GSE29378 MA", parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
      
# Fold change
FC <- FC[, unique(colnames(FC))]
write.table(FC, file='./FoldChange_GSE29378_MA.tsv', sep='\t', row.names=F, quote=F)
FC_OBJ <- File('./FoldChange_GSE29378_MA.tsv',name = 'Differential Expression Analysis GSE29378 MA logFC',parentId = parentId)
FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
    
# Pvalue
PVAL <- PVAL[, unique(colnames(PVAL))]
write.table(PVAL, file='./AdjustedPval_GSE29378_MA.tsv', sep='\t', row.names=F, quote=F)
PVAL_OBJ <- File('./AdjustedPval_GSE29378_MA.tsv',name = 'Differential Expression Analysis GSE29378 MA adjPval',parentId = parentId)
PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
```
|  *Results*           |  *SynapseID*                    |
|  -------             |   ---------                     |
|  Log fold change     |  `r FC_OBJ$properties$id`       |
|  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |